Quantum-assisted unsupervised data clustering

### 2020-05-06 manuscript.tex
Fix latex style and pictures

On a step $t$ when the BMU $w_{c}$ is selected, 
the weights of the BMU and its neigbours on feature map are adjusted according to
%
\begin{equation}
    \label{eq:learning}
	\vec{w_{i}}(t + 1) 
	= \vec{w_i}(t)
	+ \theta(c, i, t) \alpha(t) 
		\left(\vec{x}(t) - \vec{w_i}(t)\right),
\end{equation}
%
where  $\alpha(t)$ is the learning rate and $\theta(c, i, t)$ is the neighbor function,
which defines the topological neighbor neurons to be updated. 
Note that neighbor function depend on distance on feature map given initialy, 
but not distance metrics between vectors.

Intuitively, this procedure can be geometrically interpreted as iteratively moving the cluster vectors in space one at a time in a way 
that ensures each move is following the current trends inferred from their distances to the input objects. 
A visualisation of this process is shown in Fig. \ref{fig:sofm_fitting}.

@@

Let us assume we have $k$ input vectors and $l$ cluster states. 
The $i$th input vector and $j$th cluster vector are respectively denoted as $\left| x_i \right\rangle$, $\left| y_j \right\rangle$. 
The registers $\left| X \right\rangle$ and $\left| Y \right\rangle$ are initialized to store the input vectors and cluster vectors according to
%
\begin{align}
    \left| X \right\rangle  & = \frac{1}{\sqrt{k}} \sum\limits_{i=1}^{k} \left| x_i \right\rangle,  \\
    \left| Y \right\rangle&  = \frac{1}{\sqrt{l}} \sum\limits_{j=1}^{l} \left| y_j \right\rangle .
    \label{eq:encodnig}
\end{align}
% 
The two registers along with the auxiliary qubit comprise the initial state of the quantum computer according to
%
\begin{equation} 
\left| \psi_0 \right\rangle = 
    \left| X \right\rangle
    \left| Y \right\rangle 
    \left| a \right\rangle
    \label{eq:initial_state}
\end{equation}
%
where $\left| a \right\rangle$ is an auxiliary qubit in the state $\left| 0 \right\rangle$ initially.

@@

\begin{equation}
    | \psi_1 \rangle  =  
    \frac{1}{\sqrt{kl}} \sum_{i, j=1}^{k} 
    | d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \rangle 
    | y^{(1)}_j, \dots, y^{(n)}_j \rangle
    | 0 \rangle 
\end{equation}
%
where $d^{(\alpha)}_{ij} = \mathrm{CNOT}(y^{(\alpha)}_i, x^{(\alpha)}_j)$, and $\alpha = 1 \dots n$  is the qubit index in the register. 
At this stage of the computation the $\left| X \right\rangle$ no longer stores the input vectors,
instead it stores the information about pairwise different qubits between the input vector $\{X\}$ and cluster vector $\{Y\}$. 
Next, for each pair $\{X\}$ and $\{Y\}$, the accumulated information of all the differences is projected onto the amplitude of the superposed state. 
This is achieved by applying the Hadamard gate on auxiliary qubit, 
followed by a controlled rotation around z-axis gate~(\ref{eq:controled_rotation}) on $\left|a d_{ij}^{(\alpha)}\right\rangle$ 
where $d_{ij}^{(\alpha)}$ is the control qubit and $\left| a \right\rangle$ ancilla qubit register is the target.
%
\begin{equation}
    \label{eq:controled_rotation}
    C_{R_z}(\phi) = 
    \begin{pmatrix}
        1 & 0 & 0 & 0 \\
        0 & e^{-i \frac \phi 2} & 0 & 0 \\
        0 & 0 & 1 & 0 \\
		0 & 0 & 0 & e^{i \frac \phi 2}
    \end{pmatrix} .
    \quad \phi = \frac{\pi}{n}
\end{equation}
%
Finally, another Hadamard gate is applied on the ancilla qubit (see Fig. \ref{fig:qcircuit}).

After the first Hadamard on the ancilla qubit the state is
%
\begin{equation}
    \left| \psi_2 \right\rangle = 
    \frac{1}{\sqrt{kl}}\sum\limits_{i, j=1}^{k} 
    \left| d_{ij} \right\rangle 
    \left| y_j \right\rangle
    \dfrac{(\left| 0 \right\rangle + \left| 1 \right\rangle)}{\sqrt{2}}  .
\end{equation}
%
Applying the controlled rotation around $z$-axis gate on $\left| x^{(\alpha)} a \right\rangle$ where $\alpha = 1\dots n$ and the state then becomes
%
\begin{multline}
    \left| \psi_3 \right\rangle  
	= \dfrac{1}{\sqrt{2kl}} \sum\limits_{i, j=1}^{k} 
		\exp\left(
		    \dfrac{-i \pi}{2n}
		    \sum\limits_{\alpha=1}^n d^{(\alpha)}_{ij}
		\right)
		\left| d_{ij} \right\rangle 
		\left| y_j \right\rangle 
		\left| 0 \right\rangle \\
    + \dfrac{1}{\sqrt{2kl}} \sum\limits_{i, j=1}^{k}
		\exp\left(
		    \dfrac{i \pi}{2n}
		    \sum\limits_{\alpha=1}^n d^{(\alpha)}_{ij} 
		\right)
		\left| d_{ij} \right\rangle  
		\left| y_j \right\rangle 
		\left| 1 \right\rangle
\end{multline}
%
Applying another Hadamard on the ancilla qubit we obtain
%
\begin{multline}
    \left| \psi_4 \right\rangle = 
    \frac{1}{\sqrt{kl}}\sum\limits_{i, j=1}^{k} 
        \left[ 
			\cos\left(
			    \dfrac{\pi}{2n}
			    \sum\limits_{\alpha=1}^n d^{(\alpha)}_{ij}
			\right)
			\left| d_{ij} \right\rangle 
			\left| y_j \right\rangle 
			\left| 0 \right\rangle\right.
			\\+ 
			\left. i \sin\left(
			    \dfrac{\pi}{2n} 
			    \sum\limits_{\alpha=1}^n d^{(\alpha)}_{ij} 
			\right)
			\left| d_{ij} \right\rangle 
			\left| y_j \right\rangle 
			\left| 1 \right\rangle
		\right] .
\end{multline}
%
This completes the step for projecting differences between pairs of $\{X\}$ and $\{Y\}$ onto the amplitude of the auxiliary qubit. 
The process is done in the $x$-basis, achieved by the surrounding Hadamard gates. 
There are two possible measurement outcomes of the auxiliary qubit.  
Each pair of $\{X\}$ and $\{Y\}$ forms a subspace of the Hilbert space, 
the controlled rotation around $z$-axis gate ensures to change amplitudes of those outcomes within this subspace depending on how different the spin configurations between $\{X\}$ and $\{Y\}$ are.

@@

From the statistics of the measurement outcomes of final state~(\ref{eq:final_state}) we recreate the amplitudes of ancilla qubit states. 
From those amplitudes estimations we are able to plot the distance matrix between two data sets of binary vectors.
The probability amplitude of the ancilla qubit outcomes captures the exact Hamming distance as the result of the preprocessing function.
There are two possible outcomes of measurement of the ancilla qubit, each has own probability amplitude and own interpretation of that amplitude. 
For instance, for the \left| 0 \right\rangle outcome, the larger the amplitude the smaller the Hamming distance, 
and for the \left| 1 \right\rangle outcome it is the other way around, magnitude of the amplitude of that outcome is proportional to the Hamming distance.

@@

Because of that property, those algorithms fall into a special case category for which, as there is only one input vector considered, the ``Decoding'' stage (Fig.\ref{fig:qcircuit}) can be removed as measurement no longer needs to indicate for which input vector the distance has been measured.
In this special case scenario the circuit depth complexity is matching with \cite{shuld2014}.

@@

In classical SOFM \cite{kohonen1990}, first a distance calculation between a a sample vector and all cluster vectors is made, 
then the closest cluster vector is shifted towards the sample vector. 
The complexity of algorithm, in the sense of the number of distance calculations, scales as $O(LMN)$, 
where $N$ is number of samples, 
$M$ is number of randomly sampled cluster vectors, 
and $L$ is number of the shifts of cluster vectors.  
In the QASOFM described in the previous section, distance calculations are realized on a quantum device (i.e. the IBM Q Experience) with the use of circuit presented in Fig. \ref{fig:qcircuit}. This approach allows one to reduce the number of operations in a number of cluster states with an optimized number of gates 
that are possible to realize on  quantum computing devices that are currently available. 
The circuit realized in such a manner 
that calculation of Hamming distance between the sample vector and all cluster vectors is realized in one operation. 
The complexity of the quantum assisted SOFM then scales as $O(LN)$.

In order to check that our algorithm gives the expected results we compare it to classical calculations of the distance matrix on two data sets of binary vectors, as shown in Fig. \ref{fig:distance_matrix}.  
We see good agreement between the distance matrices calculated classically and on the IBM Q Experience. 
The theoretical calculations and classical simulations show perfect agreement with each other (Fig. \ref{fig:distance_matrix}(a)). 
The difference between Figs. \ref{fig:distance_matrix} (a) and (b) appears due to noise in the currently available non fault-tolerant quantum processors.

An example of the QASOFM learning process is shown in Fig.~\ref{convergence}. 
Initially, the cluster vectors were randomly chosen (see Fig.~\ref{convergence}) 
and the label of sample distribution is shown for the zeroth epoch in Fig.~\ref{convergence}(a). In order to prepare a superposition of cluster vectors needed for the calculation of the distance matrix we use the standard initialization of QISKIT library. 
Each epoch of the algorithm consists of distance calculation between all data and cluster vectors 
and requires 9 distance calculations in the  quantum implementation ($N$ in general case) 
or 27 distance calculations for classical realization 
($MN$ in general case). 
After the distance calculation from each sample to all cluster vectors at each epoch we label each sample with the index of the closest cluster vector 
and shift the closest cluster vectors to the sample one. 
The shift is made by the change of the first binary element in the cluster vectors different from the sample one. 
The evolution of the labels presented on Fig. ~\ref{convergence}(c).  
Good convergence is already observed in the fourth epoch.

@@

%\begin{thebibliography}{99}
%\end{thebibliography}
\end{document}

>>>>

On a step $t$ when the BMU $\vec{w}_{c}$ for a input $\vec{x}(t)$ is selected, 
the weights $\vec{w}_{i}$ of the BMU and its neigbours on feature map are adjusted according to
%
\begin{equation}
    \label{eq:learning}
	\vec{w}_{i}(t + 1) 
	= \vec{w}_{i}(t)
	+ \theta(c, i, t) \alpha(t) 
		\left(\vec{x}(t) - \vec{w}_{i}(t)\right) ,
\end{equation}
%
where  $\alpha(t)$ is the learning rate and $\theta(c, i, t)$ is the neighbor function,
which defines the topological neighbor neurons to be updated. 
Note that neighbor function depend on distance on feature map given initialy, 
but not distance metrics between vectors.

Intuitively, this procedure can be geometrically interpreted as iteratively moving the cluster vectors in space one at a time in a way 
that ensures each move is following the current trends inferred from their distances to the input objects. 
A visualisation of this process is shown in Fig.~\ref{fig:sofm_fitting}.

@@

Let us assume we have $k$ input vectors and $l$ cluster states. 
The $i$th input vector and $j$th cluster vector are respectively denoted as $\left| x_i \right\rangle$, $\left| y_j \right\rangle$. 
The registers $\left| X \right\rangle$ and $\left| Y \right\rangle$ are initialized to store the input vectors and cluster vectors according to
%
\begin{align}
    \label{eq:encodnig}
    \left| X \right\rangle  & = \frac{1}{\sqrt{k}} \sum\limits_{i=1}^{k} \left| x_i \right\rangle,  \\
    \left| Y \right\rangle&  = \frac{1}{\sqrt{l}} \sum\limits_{j=1}^{l} \left| y_j \right\rangle .
\end{align}
% 
The two registers along with the auxiliary qubit comprise the initial state of the quantum computer according to
%
\begin{equation} 
    \label{eq:initial_state}
	\left| \psi_0 \right\rangle = 
    \left| X \right\rangle
    \left| Y \right\rangle 
    \left| a \right\rangle ,
\end{equation}
%
where $\left| a \right\rangle$ is an auxiliary qubit in the state $\left| 0 \right\rangle$ initially.

@@

\begin{equation}
    | \psi_1 \rangle  =  
    \frac{1}{\sqrt{kl}} \sum_{i, j=1}^{k} 
    | d^{(1)}_{ij}, \dots, d^{(n)}_{ij} \rangle 
    | y^{(1)}_j, \dots, y^{(n)}_j \rangle
    | 0 \rangle ,
\end{equation}
%
where $d^{(\alpha)}_{ij} = \mathrm{CNOT}(y^{(\alpha)}_i, x^{(\alpha)}_j)$, and $\alpha = 1 \dots n$  is the qubit index in the register. 
At this stage of the computation the $\left| X \right\rangle$ no longer stores the input vectors,
instead it stores the information about pairwise different qubits between the input vector $\{X\}$ and cluster vector $\{Y\}$. 
Next, for each pair $\{X\}$ and $\{Y\}$, the accumulated information of all the differences is projected onto the amplitude of the superposed state. 
This is achieved by applying the Hadamard gate on auxiliary qubit, 
followed by a controlled rotation around z-axis gate~(\ref{eq:controled_rotation}) on $\left|a d_{ij}^{(\alpha)}\right\rangle$ 
where $d_{ij}^{(\alpha)}$ is the control qubit and $\left| a \right\rangle$ ancilla qubit register is the target.
%
\begin{equation}
    \label{eq:controled_rotation}
    C_{R_z}(\phi) = 
    \begin{pmatrix}
        1 & 0 & 0 & 0 \\
        0 & e^{-i \frac \phi 2} & 0 & 0 \\
        0 & 0 & 1 & 0 \\
		0 & 0 & 0 & e^{i \frac \phi 2}
    \end{pmatrix} ,
    \quad \phi = \frac{\pi}{n}.
\end{equation}
%
Finally, another Hadamard gate is applied on the ancilla qubit (see Fig.~\ref{fig:qcircuit}).

After the first Hadamard on the ancilla qubit the state is
%
\begin{equation}
    \left| \psi_2 \right\rangle = 
    \frac{1}{\sqrt{kl}}\sum\limits_{i, j=1}^{k} 
    \left| d_{ij} \right\rangle 
    \left| y_j \right\rangle
    \dfrac{(\left| 0 \right\rangle + \left| 1 \right\rangle)}{\sqrt{2}}  .
\end{equation}
%
Applying the controlled rotation around $z$-axis gate on $\left| x^{(\alpha)} a \right\rangle$ where $\alpha = 1\dots n$ and the state then becomes
%
\begin{multline}
    \left| \psi_3 \right\rangle  
	= \dfrac{1}{\sqrt{2kl}} \sum\limits_{i, j=1}^{k} 
		\exp\left(
		    \dfrac{-i \pi}{2n}
		    \sum\limits_{\alpha=1}^n d^{(\alpha)}_{ij}
		\right)
		\left| d_{ij} \right\rangle 
		\left| y_j \right\rangle 
		\left| 0 \right\rangle \\
    + \dfrac{1}{\sqrt{2kl}} \sum\limits_{i, j=1}^{k}
		\exp\left(
		    \dfrac{i \pi}{2n}
		    \sum\limits_{\alpha=1}^n d^{(\alpha)}_{ij} 
		\right)
		\left| d_{ij} \right\rangle  
		\left| y_j \right\rangle 
		\left| 1 \right\rangle .
\end{multline}
%
Applying another Hadamard on the ancilla qubit we obtain
%
\begin{multline}
    \left| \psi_4 \right\rangle = 
    \frac{1}{\sqrt{kl}}\sum\limits_{i, j=1}^{k} 
        \left[ 
			\cos\left(
			    \dfrac{\pi}{2n}
			    \sum\limits_{\alpha=1}^n d^{(\alpha)}_{ij}
			\right)
			\left| d_{ij} \right\rangle 
			\left| y_j \right\rangle 
			\left| 0 \right\rangle\right.
			\\+ 
			\left. i \sin\left(
			    \dfrac{\pi}{2n} 
			    \sum\limits_{\alpha=1}^n d^{(\alpha)}_{ij} 
			\right)
			\left| d_{ij} \right\rangle 
			\left| y_j \right\rangle 
			\left| 1 \right\rangle
		\right] .
\end{multline}
%
This completes the step for projecting differences between pairs of $\{X\}$ and $\{Y\}$ onto the amplitude of the auxiliary qubit. 
The process is done in the $x$-basis, achieved by the surrounding Hadamard gates. 
There are two possible measurement outcomes of the auxiliary qubit.  
Each pair of $\{X\}$ and $\{Y\}$ forms a subspace of the Hilbert space, 
the controlled rotation around $z$-axis gate ensures to change amplitudes of those outcomes within this subspace depending on how different the spin configurations between $\{X\}$ and $\{Y\}$ are.

@@

From the statistics of the measurement outcomes of final state~(\ref{eq:final_state}) we recreate the amplitudes of ancilla qubit states. 
From those amplitudes estimations we are able to plot the distance matrix between two data sets of binary vectors.
The probability amplitude of the ancilla qubit outcomes captures the exact Hamming distance as the result of the preprocessing function.
There are two possible outcomes of measurement of the ancilla qubit, each has own probability amplitude and own interpretation of that amplitude. 
For instance, for the $\left| 0 \right\rangle$ outcome, the larger the amplitude the smaller the Hamming distance, 
and for the $\left| 1 \right\rangle$ outcome it is the other way around, magnitude of the amplitude of that outcome is proportional to the Hamming distance.

@@

Because of that property, those algorithms fall into a special case category for which, as there is only one input vector considered, the ``Decoding'' stage (Fig.~\ref{fig:qcircuit}) can be removed as measurement no longer needs to indicate for which input vector the distance has been measured.
In this special case scenario the circuit depth complexity is matching with \cite{schuld2014}.

@@

In classical SOFM \cite{kohonen1990}, first a distance calculation between a a sample vector and all cluster vectors is made, 
then the closest cluster vector is shifted towards the sample vector. 
The complexity of algorithm, in the sense of the number of distance calculations, scales as $O(LMN)$, 
where $N$ is number of samples, 
$M$ is number of randomly sampled cluster vectors, 
and $L$ is number of the shifts of cluster vectors.  
In the QASOFM described in the previous section, distance calculations are realized on a quantum device (i.e. the IBM Q Experience) with the use of circuit presented in Fig.~\ref{fig:qcircuit}. This approach allows one to reduce the number of operations in a number of cluster states with an optimized number of gates 
that are possible to realize on  quantum computing devices that are currently available. 
The circuit realized in such a manner 
that calculation of Hamming distance between the sample vector and all cluster vectors is realized in one operation. 
The complexity of the quantum assisted SOFM then scales as $O(LN)$.

In order to check that our algorithm gives the expected results we compare it to classical calculations of the distance matrix on two data sets of binary vectors, as shown in Fig.~\ref{fig:distance_matrix}.  
We see good agreement between the distance matrices calculated classically and on the IBM Q Experience. 
The theoretical calculations and classical simulations show perfect agreement with each other (Fig.~\ref{fig:distance_matrix}(a)). 
The difference between Figs.~\ref{fig:distance_matrix} (a) and (b) appears due to noise in the currently available non fault-tolerant quantum processors.

An example of the QASOFM learning process is shown in Fig.~\ref{convergence}. 
Initially, the cluster vectors were randomly chosen (see Fig.~\ref{convergence}) 
and the label of sample distribution is shown for the zeroth epoch in Fig.~\ref{convergence}(a). In order to prepare a superposition of cluster vectors needed for the calculation of the distance matrix we use the standard initialization of QISKIT\cite{qiskit} library. 
Each epoch of the algorithm consists of distance calculation between all data and cluster vectors 
and requires 9 distance calculations in the  quantum implementation ($N$ in general case) 
or 27 distance calculations for classical realization 
($MN$ in general case). 
After the distance calculation from each sample to all cluster vectors at each epoch we label each sample with the index of the closest cluster vector 
and shift the closest cluster vectors to the sample one. 
The shift is made by the change of the first binary element in the cluster vectors different from the sample one. 
The evolution of the labels presented on Fig.~\ref{convergence}(c).  
Good convergence is already observed in the fourth epoch.

@@

\end{document}
